---
title: "Assignment IV"
subtitle: "DATA471"
author: "Name: Hamid Mehmood        Student ID: 300693052"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
    df_print: kable
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4.5,
                      fig.pos = "H", out.extra = "")
set.seed(301471)
```

# Exercise 1

## Part 1: Which sample quantiles should we use? *(10%)*

To compute the IQR estimator, $\hat\lambda_{\text{IQR}} = \ln 3 / (q_{3/4}-q_{1/4})$, I need
the first and third quartiles of my sample. That sounds like a non-issue, but it turns out
there is no single "correct" way to pull a quartile out of a finite sample. Wicklin (2017)
goes through nine different definitions that all show up in real software, and they can
genuinely disagree on the same data. So before I trust my quartiles, I should know which
definition I am actually using.

### Summary of Wicklin (2017)

Wicklin's post is basically a friendly walk through the nine quantile definitions that
Hyndman and Fan (1996) catalogued (I will call them H and F). Here is what I took from it.

**The basic idea.** Sort the sample $x_{(1)}\le\dots\le x_{(N)}$. To estimate the $p$-th
quantile you find the two neighbouring order statistics that sit either side of it, and then
decide what to do in the gap between them. Every one of the nine methods can be written in
the same general form,
$$
q=(1-\lambda)\,x_{(j)}+\lambda\,x_{(j+1)},\qquad j=\lfloor Np+m\rfloor,
$$
where $\lambda\in[0,1]$ is an interpolation weight and $m$ is a small constant that changes
with the method. So the nine "definitions" are not nine unrelated ideas; they are nine
choices of $m$ and of how you set $\lambda$.

**Two families.** Wicklin splits them into two groups. Types 1 to 3 are *discrete* (rounding)
methods: the answer is always an actual data value (or the average of two), so the quantile
function looks like a staircase. Types 4 to 9 are *continuous* (interpolation) methods: they
join the points up with straight lines, so the answer can be anything between two
observations.

**Where they disagree.** The methods differ most for small samples and out in the tails (or
wherever there is a big gap between neighbouring values). Wicklin shows that for a sample of
100 the nine methods basically land on top of each other. His honest summary is "yes and no":
the choice matters in theory and for tiny samples, but hardly at all for large ones. Either
way, his advice is to always say which method you used.

**Software defaults.** This is the part that catches people out. SAS supports five of the
nine (the `QNTLDEF=` option) and defaults to type 2, a discrete method, while R defaults to
type 7, a continuous one. Same data, different quartiles.

### Which definition would I use, and why?

I will use a continuous method throughout, specifically **type 8**. My reasons, in plain
terms:

1. My estimator depends on a *spread*, $q_{3/4}-q_{1/4}$, not on a single order statistic.
   Interpolation uses both of the bracketing points, so it gives a smoother spread than the
   discrete methods, which can snap both quartiles onto the same repeated value.
2. Among the continuous methods, type 8 is the one Hyndman and Fan actually recommend,
   because it is approximately median-unbiased whatever the underlying distribution is. Since
   the very next part of this exercise is about bias, it feels right to start from the
   quartile rule with the best bias behaviour.
3. Being honest, for $n=100$ the choice barely matters, but for the $n=5$ and $n=30$ cases
   coming up it really does, so it is worth picking on purpose rather than by accident.

In R this is just `quantile(x, type = 8)`. Type 7 (R's default) would be fine too and would
not change much for larger $n$.

### What is the default in R and Python?

- **R:** `quantile()` defaults to **type 7** (continuous, linear interpolation). You change
  it with the `type` argument.
- **Python:** `numpy` and `pandas` also default to linear interpolation, which is **the same
  as R's type 7** (`method = "linear"` in current NumPy).
- **SAS (for contrast):** defaults to **type 2**, a discrete method, which is exactly why SAS
  and R/Python can give different quartiles on a small sample.

## Part 2: A bootstrap bias-corrected IQR estimator *(20%)*

### Deriving $\tilde\lambda$

The bias of the IQR estimator is
$$
\text{Bias}(\hat\lambda_{\text{IQR}}) \;=\; \mathbb E_\lambda\!\big[\hat\lambda_{\text{IQR}}(\mathbf X)\big]-\lambda .
$$
I cannot just compute this, for two reasons: it depends on the true $\lambda$ (which is the
thing I am trying to estimate), and the sampling distribution of a ratio of quartiles has no
nice closed form. This is exactly the situation the bootstrap is built for.

The bootstrap idea is to treat my observed sample as if it were the whole population, and to
draw new samples from it by resampling **with replacement**. Inside this "bootstrap world" I
know everything, so I can measure the bias directly. The value I actually computed,
$\hat\lambda_{\text{IQR}}(\mathbf x)$, plays the role of the truth, and I estimate the bias by
Monte Carlo:
$$
\widehat{\text{Bias}}
= \underbrace{\frac1B\sum_{b=1}^{B}\hat\lambda_{\text{IQR}}(\mathbf X^{*}_b)}_{\text{mean over }B\text{ resamples}}
\;-\; \hat\lambda_{\text{IQR}}(\mathbf x),
\qquad \mathbf X^{*}_b \text{ drawn with replacement from } \mathbf x .
$$
The first-order bias-corrected estimator just subtracts this estimated bias:
$$
\boxed{\;\tilde\lambda
= \hat\lambda_{\text{IQR}}(\mathbf x)-\widehat{\text{Bias}}
= 2\,\hat\lambda_{\text{IQR}}(\mathbf x)-\frac1B\sum_{b=1}^{B}\hat\lambda_{\text{IQR}}(\mathbf X^{*}_b)\;}
$$
As instructed, I fix $B=200$ resamples per input sample.

### Implementation

Here are the three estimators. I use R's built-in `quantile(type = 8)` throughout, so the
code stays as close to the formulas as possible.

```{r estimators}
lambda_ml  <- function(x) 1 / mean(x)

lambda_iqr <- function(x) {
  qs <- quantile(x, probs = c(0.25, 0.75), type = 8, names = FALSE)
  log(3) / (qs[2] - qs[1])
}

lambda_tilde <- function(x, B = 200) {
  n     <- length(x)
  lhat  <- lambda_iqr(x)
  boot  <- replicate(B, lambda_iqr(sample(x, n, replace = TRUE)))
  fail  <- sum(!is.finite(boot))            # resamples whose IQR came out zero
  boot  <- boot[is.finite(boot)]
  list(est  = 2 * lhat - mean(boot),        # = lhat - estimated bias
       fail = fail)
}
```

The `fail` value in that return is there for a concrete reason. When $n$ is very small, a
bootstrap resample can repeat values so that its first and third quartiles come out equal,
which gives a zero IQR and an infinite estimate. The `is.finite` line drops those degenerate
resamples before averaging so that $\tilde\lambda$ stays finite, and `fail` records how many
had to be dropped. That count is essentially zero for moderate samples but noticeable at
$n=5$, so keeping it gives me a real handle on why the estimator becomes unreliable for tiny
samples, which I return to in Part 3.

As a quick check that the correction pushes in the right direction: the ML estimator of an
exponential rate has a known bias, $\mathbb E[\hat\lambda_{\text{ML}}]=\frac{n}{n-1}\lambda$,
so it overestimates for small $n$, and the IQR estimator overestimates too. A good bias
correction should pull $\tilde\lambda$ back down towards $\lambda$ on average, which is what
happens in Part 3.

## Part 3: Monte Carlo comparison *(20%)*

### Design

I compare the three estimators on the grid $\Lambda\times N$ with $\lambda\in\{0.1,1,10\}$
and $n\in\{5,30,100\}$, using **mean squared error** as the only metric,
$\text{MSE}(\hat\lambda)=\mathbb E[(\hat\lambda-\lambda)^2]$, estimated by averaging
$(\hat\lambda-\lambda)^2$ over $R$ independent replications.

### How many replications, and why $\lambda$ barely matters

There is a nice structural fact that basically decides this for me. All three estimators are
**scale-equivariant**: if I rescale the data $x\mapsto cx$, the estimate becomes
$\hat\lambda/c$ (true for the mean, for the quartile gap, and so for $\tilde\lambda$ too).
Since $X\sim\text{Exp}(\lambda)$ means $\lambda X\sim\text{Exp}(1)$, the distribution of the
relative error $\hat\lambda/\lambda$ does not depend on $\lambda$ at all. So
$$
\text{MSE}(\hat\lambda;\lambda)=\lambda^{2}\,\text{MSE}(\hat\lambda;1),
$$
the MSE is just $\lambda^2$ times a number that depends only on $n$ and the estimator.

Two things follow. First, I do not need more replications for different $\lambda$; the same
$R$ gives the same relative precision everywhere. Second, I can use this as a free check: the
columns of $\text{MSE}/\lambda^2$ should match across $\lambda$.

For the size of $R$, I follow Robert and Casella (2009) and treat the estimated MSE as a
sample mean, watching its Monte Carlo standard error,
$\text{SE}=\text{sd}\big((\hat\lambda-\lambda)^2\big)/\sqrt R$. I use $R=3000$ for every cell.
That gives standard errors of only a couple of percent of the MSE for $n=30$ and $n=100$,
which is plenty. The one exception is $n=5$ for the IQR-based estimators, whose squared error
is very heavy-tailed. There, no realistic $R$ settles the estimate down, and the honest
takeaway is not "run more replications" but "do not use these estimators at $n=5$" (more on
that below).

### Running the experiment

The loop below calls the three estimators directly. Each $\tilde\lambda$ does 200 bootstrap
resamples, so the whole run takes a few minutes; the chunk is cached so re-knitting is
instant.

```{r simulation, cache=TRUE}
Lambda <- c(0.1, 1, 10); Ns <- c(5, 30, 100); R <- 3000
set.seed(301471)

rows <- list(); k <- 0
for (lam in Lambda) for (n in Ns) {
  ml <- iqr <- til <- fails <- numeric(R)
  for (r in 1:R) {
    x <- rexp(n, rate = lam)
    ml[r]  <- lambda_ml(x)
    iqr[r] <- lambda_iqr(x)
    tt <- lambda_tilde(x); til[r] <- tt$est; fails[r] <- tt$fail
  }
  mse <- function(e) mean((e - lam)^2)
  se  <- function(e) sd((e - lam)^2) / sqrt(R)
  k <- k + 1
  rows[[k]] <- data.frame(
    lambda = lam, n = n,
    MSE_ML = mse(ml), MSE_IQR = mse(iqr), MSE_TIL = mse(til),
    seML = se(ml), seIQR = se(iqr), seTIL = se(til),
    biasML = mean(ml) - lam, biasIQR = mean(iqr) - lam, biasTIL = mean(til) - lam,
    varIQR = var(iqr), varTIL = var(til), meanfail = mean(fails))
}
res <- do.call(rbind, rows)
```

### Results

```{r table-mse}
tab <- data.frame(
  lambda = res$lambda, n = res$n,
  MSE_ML  = signif(res$MSE_ML, 3),
  MSE_IQR = signif(res$MSE_IQR, 3),
  MSE_tilde = signif(res$MSE_TIL, 3))
knitr::kable(tab, caption = "MSE of the three estimators over the grid (R = 3000).")
```

```{r table-relative}
rel <- data.frame(
  lambda = res$lambda, n = res$n,
  rel_ML  = signif(res$MSE_ML  / res$lambda^2, 3),
  rel_IQR = signif(res$MSE_IQR / res$lambda^2, 3),
  rel_tilde = signif(res$MSE_TIL / res$lambda^2, 3))
knitr::kable(rel, caption = "Relative MSE = MSE / $\\lambda^2$. For each n these should agree across lambda (scale-equivariance), and they do for n = 30 and 100. The n = 5 tilde column does not, which reveals its instability.")
```

```{r plot, fig.cap="Relative MSE (MSE / $\\lambda^2$, averaged over the three lambda) versus n on a log scale. Lower is better. ML is best, the bootstrap correction sits just above plain IQR, and the tilde estimator at n = 5 is off the charts."}
agg <- aggregate(cbind(MSE_ML/lambda^2, MSE_IQR/lambda^2, MSE_TIL/lambda^2) ~ n,
                 data = res, FUN = mean)
names(agg) <- c("n", "ML", "IQR", "tilde")
matplot(agg$n, agg[,-1], type = "b", log = "y", pch = 19, lty = 1,
        col = c("#1b7837", "#2166ac", "#b2182b"),
        xlab = "sample size n", ylab = "relative MSE (log scale)",
        main = "Relative MSE vs n")
legend("topright", legend = c("ML", "IQR", "tilde (bias-corrected)"),
       col = c("#1b7837", "#2166ac", "#b2182b"), pch = 19, lty = 1, bty = "n")
```

```{r table-bias}
bt <- data.frame(
  lambda = res$lambda, n = res$n,
  bias_IQR = signif(res$biasIQR, 3),
  bias_tilde = signif(res$biasTIL, 3),
  var_IQR = signif(res$varIQR, 3),
  var_tilde = signif(res$varTIL, 3),
  avg_fails = signif(res$meanfail, 3))
knitr::kable(bt, caption = "Bias and variance of the IQR estimator vs its bias-corrected version, plus the average number of degenerate (zero-IQR) bootstrap resamples per sample (out of B = 200).")
```

### What the experiment tells us

**1. ML wins on MSE everywhere.** Since the exponential model is correct for the simulated
data, this is exactly what I would expect: ML uses the whole sample and is (asymptotically)
efficient, while the IQR estimator throws away everything except two quartiles. Roughly, the
IQR estimator carries about 2 to 2.5 times the MSE of ML for $n=30$ and $n=100$.

**2. Scale-equivariance holds, a nice built-in check.** For $n=30$ and $n=100$ the
relative-MSE table is basically constant across $\lambda$ (for example ML is about 0.040 at
$n=30$ and about 0.011 at $n=100$ for every $\lambda$). That both reassures me the code is
right and backs up using the same $R$ for all $\lambda$.

**3. The bootstrap correction reduces bias but not MSE.** This is the most interesting bit.
Comparing the plain IQR estimator with $\tilde\lambda$ in the bias table, the correction does
shrink the bias: for $n=100,\ \lambda=1$ it moves from about $+0.016$ to about $-0.008$, even
flipping sign, which is the slight overshoot that first-order corrections are known for. But
the variance goes up (each bootstrap average adds its own noise), and since
$\text{MSE}=\text{bias}^2+\text{variance}$ with variance dominating here, $\tilde\lambda$
actually ends up with a slightly larger MSE than the plain IQR estimator. It is a clean
example of the bias-variance trade-off: fixing the bias is not free.

**4. $n=5$ is a cautionary tale.** At the smallest sample size everything is poor, but the
IQR-based estimators are genuinely unreliable. Because a resample is drawn from only five
points, the bootstrap frequently hits ties or near-ties, which throw up huge estimates and
give $\tilde\lambda$ an essentially heavy-tailed, near-infinite-variance distribution. The
failure column in the table above makes this concrete: at $n=5$ a nonzero number of resamples
per sample collapse to a zero IQR, while at $n=30$ and $n=100$ essentially none do. Its
estimated MSE at $n=5$ jumps around a lot and its Monte Carlo standard error is about the same
size as the estimate itself, so those particular numbers should be read as "very large and
unstable" rather than taken literally. The practical message: the IQR estimator, and
especially its bootstrap correction, needs a decent sample size (around $n=30$ or more) to
behave.

**Bottom line.** If I trust the exponential model, I would use the ML estimator. The IQR
estimator is a reasonable, simple backup for moderate $n$, and it would really earn its keep
if I were worried about outliers or contamination (which is the whole point of a robust
estimator, though I have not tested that here). The bootstrap bias correction does what it
promises for bias, but it is not a free win for MSE, and it should not be used at very small
$n$.

# Exercise 2

```{r load-data, message=FALSE}
library(fpp2)   # the ausbeer dataset comes from this package
```

## The dataset

The `ausbeer` series is the quarterly beer production of Australia, measured in megalitres.
It is one of the example series Hyndman and Athanasopoulos (2018) use in their textbook, so
it is a good series to practise looking at before doing any modelling. Everything below is
plain exploratory data analysis: I look at the series, describe what I see, and back each
claim up with a plot.

## Q1: What is the frequency of the series?

The frequency is **4**: there are four observations per year, one per quarter. There are two
easy ways to see this.

```{r q1-meta}
frequency(ausbeer)
c(start = paste(start(ausbeer), collapse = " Q"),
  end   = paste(end(ausbeer),   collapse = " Q"))
```

First, the series is stored as a quarterly time series that runs from 1956 Q1 to 2010 Q2, so
its `frequency` attribute is 4. Second, and more importantly, you can read it straight off a
plot. If I zoom in on the first few years, the pattern repeats itself every four observations,
with a clear tall peak once a year and a dip in between.

```{r q1-zoom, fig.height=3.8, fig.cap="Zooming in on 1956 to 1962: the pattern repeats every four quarters, so the frequency is 4."}
plot(window(ausbeer, end = c(1962, 4)), ylab = "megalitres", xlab = "year",
     main = "First few years of ausbeer", col = "#2166ac")
points(window(ausbeer, end = c(1962, 4)), pch = 19, cex = 0.6, col = "#2166ac")
```

Counting the points between one peak and the next gives four, which is the frequency.

## Q2: Overall behaviour and long-term trend

```{r q2-timeplot, fig.cap="The full ausbeer series. A strong rise into the mid-1970s, a long plateau, then a gradual decline."}
plot(ausbeer, ylab = "megalitres", xlab = "year",
     main = "Quarterly Australian beer production, 1956 to 2010", col = "#2166ac")
```

Yes, there is a clear long-term trend, but it is not a single straight line. Production rises
steeply from about 1956 to the mid-1970s, roughly doubling from around 250 to over 500
megalitres. It then flattens into a long plateau through the late 1970s, 1980s and early
1990s, with the highest peaks of the whole series in this stretch (up to 599). After the
mid-1990s it drifts gently downwards to the end of the series. So the trend goes up, levels
off, and then comes down again. On top of this slow movement there is a very regular
saw-tooth wobble, which is the seasonal pattern I look at next.

## Q3: Is there seasonality?

Yes, and it is strong and very regular. The quickest way to summarise it is the average
production in each quarter across all years:

```{r q3-means}
round(tapply(ausbeer, cycle(ausbeer), mean), 1)  # 1 = Q1, 2 = Q2, 3 = Q3, 4 = Q4
```

The pattern within a year is: production is high in Q1, drops to its lowest in Q2, recovers a
little in Q3, and then jumps to its highest in Q4. In other words Q4 is the big quarter and Q2
is the quiet one. This makes sense for Australia, where Q4 covers the run-up to the warm
Christmas and summer season in the Southern hemisphere. The seasonal swing is large, roughly
100 megalitres between the low and high quarters, which is not small next to a level of around
400 to 500. I look at this seasonal shape in detail in Q6 and Q7.

## Q4: Is the variability constant? Would a transformation help?

To judge the variability I plotted the size of the seasonal swing (the largest minus the
smallest value within each year) over time.

```{r q4-amplitude, fig.cap="Within-year range (max minus min) over time. Larger in the high-production decades, smaller when the level is low."}
yrs <- unique(floor(time(ausbeer)))
m <- matrix(NA, length(yrs), 4, dimnames = list(yrs, c("Q1","Q2","Q3","Q4")))
for (i in seq_along(ausbeer))
  m[as.character(floor(time(ausbeer)[i])), cycle(ausbeer)[i]] <- ausbeer[i]
full <- m[rowSums(is.na(m)) == 0, ]
amp  <- apply(full, 1, max) - apply(full, 1, min)
plot(as.numeric(rownames(full)), amp, type = "b", pch = 19, col = "#b2182b",
     xlab = "year", ylab = "within-year range", main = "Seasonal amplitude over time")
```

The variability is not perfectly constant, but the story is more subtle than "it grows with
the level". The swing is around 90 in the low-production 1950s, grows to about 150 to 175 in
the high-production 1970s to early 1990s, and then shrinks back to roughly 65 to 105 in the
2000s. The interesting part is the end: production in the 2000s is still fairly high (around
400 to 490), yet the swings are as small as they were at the very start. So the amplitude
tracks the era more than it tracks the level. A variance-stabilising transformation such as a
log or a Box-Cox assumes the swing grows in proportion to the level, and that assumption does
not really hold here (the recent years break it). So I would say a transformation is optional
at best, and I would not reach for a strong one. The variability is close enough to constant
that the trend and seasonality are the features worth worrying about.

## Q5: Any unusual observations?

Most of the series is extremely regular, so "unusual" here means points that sit clearly off
the smooth trend and their usual seasonal level. To find them I removed a centred moving
average (a simple descriptive smoother, not a fitted model) and the quarterly averages, and
looked at what was left.

```{r q5-outliers, fig.cap="The series with the most unusual points marked in red."}
ma   <- stats::filter(ausbeer, c(1/8, rep(1/4, 3), 1/8), sides = 2)  # trend-cycle
irr  <- (ausbeer - ma) - tapply(ausbeer - ma, cycle(ausbeer), mean, na.rm = TRUE)[cycle(ausbeer)]
lab  <- paste0(floor(time(ausbeer)), " Q", cycle(ausbeer))
top  <- order(abs(irr), decreasing = TRUE)[1:4]
data.frame(quarter = lab[top], value = as.numeric(ausbeer)[top],
           deviation = round(irr[top], 1))

plot(ausbeer, ylab = "megalitres", xlab = "year", col = "grey55",
     main = "Unusual observations")
points(time(ausbeer)[top], ausbeer[top], pch = 19, col = "#b2182b")
text(time(ausbeer)[top], ausbeer[top], lab[top], pos = 3, cex = 0.7, col = "#b2182b")
```

A few points stand out. The two highest observations of the entire series, 1988 Q4 (598) and
1990 Q4 (599), are spikes that sit noticeably above even the other strong Q4 values around
them. In the other direction, 1981 Q1 (548) is a strikingly strong first quarter, almost at
the level of a typical Q4, and 1980 Q4 (526) is an unusually weak fourth quarter, well below
the Q4 peaks on either side. They attract attention because each one departs from the very
steady trend-plus-season pattern that the rest of the series follows so closely. None of them
looks like a data error, they are just the years that wobbled a bit more than usual.

## Q6: Seasonal plot

```{r q6-seasonplot, fig.height=5, fig.cap="Seasonal plot: one line per year, quarters on the horizontal axis. Every year traces the same shape."}
matplot(1:4, t(m), type = "l", lty = 1,
        col = grey(seq(0.15, 0.85, length = nrow(m))),
        xaxt = "n", xlab = "quarter", ylab = "megalitres",
        main = "Seasonal plot (one line per year)")
axis(1, at = 1:4, labels = c("Q1", "Q2", "Q3", "Q4"))
```

The ordinary time plot shows the seasonal wobble, but the years blur together and it is hard
to compare the same quarter across years. The seasonal plot fixes this by drawing each year as
its own line across the four quarters. What it reveals is how consistent the seasonal shape
is: essentially every line follows the same path, down into Q2 and up to a peak in Q4. The
lines are stacked by height, which is just the trend moving the whole year up or down, but the
shape barely changes. It also makes an odd year easy to spot, because it would be a line that
does not follow the usual dip-then-peak track.

## Q7: Seasonal subseries plot

```{r q7-subseries, fig.height=4.8, fig.cap="Seasonal subseries plot. Each panel is one quarter across all years; the red line is that quarter's mean."}
monthplot(ausbeer, ylab = "megalitres", xlab = "quarter",
          main = "Seasonal subseries", col = "#2166ac", col.base = "red")
```

This plot splits the series into four little series, one per quarter, and draws each quarter's
mean as a red line. Reading off the red lines, **Q4 has the highest typical value and Q2 the
lowest**, with Q1 second and Q3 third (means of about 485, 417, 387 and 373 for Q4, Q1, Q3,
Q2). The seasonal pattern looks stable over time: within every panel the quarter follows the
same overall shape as the whole series (up to the mid-1970s, a plateau, then a decline), and
the ordering of the quarters never changes. The gap between the quarters is a little wider in
the high-production middle decades and a little narrower recently, which matches the amplitude
plot in Q4, but the pattern itself does not drift or flip.

## Q8: Would you detrend or transform before modelling?

Two separate questions here, and my visual answers are different for each. I am not fitting
anything, just arguing from the plots.

**Detrend / account for trend and season: yes.** The time plot in Q2 shows the average level
moving a lot over time (up, flat, then down), and the subseries plot in Q7 shows every quarter
drifting in the same way. A series whose mean level changes like this is not stationary, so
before modelling I would remove that structure, for example by taking a seasonal difference to
knock out the strong quarterly pattern and a first difference to deal with the remaining slow
movement in the level. The visual argument is simply that both the trend and the season are
large and obvious, so a sensible model has to deal with them first.

**Transform (log / Box-Cox): probably not.** The amplitude plot in Q4 is the argument. If the
variability grew clearly with the level, a log would be worth it, but here the swing does not
scale cleanly with the level (the recent years are fairly high but have small swings). The
variation is close enough to constant that a transformation would add complication for little
gain. If I did anything at all it would be a very mild Box-Cox, but I would be comfortable
leaving the series on its original scale.

In short: deal with the trend and the seasonality before modelling, but a variance transform
is not really needed for this series.

